# (PART) Part I: Data Types {-} 



```{r setup, include =FALSE}
library(reticulate)
use_condaenv("base")
```


# Basic Types

In every programming language, data is stored in different ways. Writing a program that manipulates data requires understanding all of the choices. That is why we must be concerned with the different **types** of data in our R and Python programs. Different types are suitable for different purposes.

There are similarities between Python's and R's type systems. However, there are may  differences as well. Be prepared for these differences. There are many more of them in this chapter than there were in the previous chapter!

If you're ever unsure what type a variable has, use `type()` (in Python) or `typeof()` (in R) to query it. 

Storing an individual piece of information is simple in both languages. However, while Python has scalar types, R does not draw as strong of a distinction between scalar and compound types.

## Basic Types In Python

In Python, the simplest types we frequently use are `str` (short for string), `int` (short for integer), `float` (short for floating point) and `bool` (short for Boolean). This list is not exhaustive, but these are a good collection to start thinking about. For a complete list of built-in types in Python, click [here](https://docs.python.org/3/library/stdtypes.html).

```{python, collapse = TRUE}
print(type('a'), type(1), type(1.3))
```

Strings are useful for processing text data such as names of people/places/things and messages such as texts, tweets and emails. If you are dealing with numbers, you need floating points if you have a number that might have a fractional part after its decimal; otherwise you'll need an integer. Booleans are useful for situations where you need to record whether something is true or false. They are also important to understand for control-flow in section \@ref(control-flow).

In the next section we will discuss the Numpy library. This library has a [broader collection](https://numpy.org/doc/stable/user/basics.types.html) of basic types that allows for finer control over any script you write. 


### Type Conversions in Python

We will often have to convert between types in a Python program. This is called **type conversion**, and it can be either implicitly or explicitly done. 

For example, `int`s are often implicitly converted to `float`s, so that arithmetic operations work. 
```{python, collapse = TRUE}
my_int = 1
my_float = 3.2
my_sum = my_int + my_float
print("my_int's type", type(my_int))
print("my_float's type", type(my_float))
print(my_sum)
print("my_sum's type", type(my_sum))
```

You might be disappointed if you always count on this behavior, though.
```{python, error = TRUE, collapse=TRUE}
3.2 + "3.2"
```

Explicit conversions occur when we as programmers explicitly ask Python to perform a conversion. We will do this with the functions such as `int()`, `str()`, `float()`, and `bool()`. 

```{python, collapse = TRUE}
my_date = "5/2/2021"
month_day_year = my_date.split('/')
my_year = int(month_day_year[-1]) 
print('my_year is equal to ', my_year, 'and its type is ', type(my_year))
```


## Basic Types In R

In R, the names of basic types are only slightly different. They are `logical` (instead of `bool`), `integer` (instead of `int`), `double` or `numeric` (instead of `float`)^["double" is short for "double precision floating point." In other programming languages, the programmer might choose how many decimal points of precision he or she wants.], `character` (instead of `str`), `complex` (for calculations involving imaginary numbers), and `raw` (useful for working with bytes).

```{R, collapse = TRUE}
# cat() is kind of like print()
cat(typeof('a'), typeof(1), typeof(1.3))
```

In this case R automatically upgraded `1` to a double. If you wanted to force it to be an integer, you can add a capital "L" to the end of the number.

```{R, collapse = TRUE}
# cat() is kind of like print()
cat(typeof('a'), typeof(1L), typeof(1.3))
```


### Type Conversions in R

You can explicitly and implicitly convert types in R just as you did in Python. Implicit conversion looks like this.

```{R, collapse = TRUE}
myInt = 1
myDouble = 3.2
mySum = myInt + myDouble
print(paste0("my_int's type is ", typeof(myInt)))
print(paste0("my_float's type is ", typeof(myDouble)))
print(mySum)
print(paste0("my_sum's type is ", typeof(mySum)))
```

Explicit conversion can be achieved with functions such as `as.integer`, `as.logical`, `as.double`, etc.

```{R, collapse = TRUE}
print(typeof(1))
print(typeof(as.logical(1)))
```



### R's Simplification

The basic types of R are a little different than the basic types of Python. On the one hand, Python has basic types for individual elements, and it uses separate types as containers for storing many elements. On the other, R uses the same type to store a single element as it does to store many elements. Strictly speaking, R does not have a scalar type. 

Technically, all of the examples we just did in R are using length one **vectors**--`logical` `integer` `double`, `character`, `complex`, and `raw` are the possible **modes** of a vector. `vector`s will be discussed further section \@ref(r-vectors-versus-numpy-arrays-and-pandas-series).

Think about which option you prefer. What are the benefits of using separate types for scalars and collections? What are the benefits of using the same type?



## Exercises

### Easier R Questions

1. Which R base type is ideal for each piece of data? Assign your answers to a `character` `vector` of length four called `questionOne`. 

    + An individual's IP address
    + whether or not an individual attended a study
    + the number of seeds found in a plant
    + the amount of time it takes for a car to race around a track
  
2. Floating points are weird. What gets printed is not the same as what is stored! In R, you can control how many digits get printed by using the `options` function. 

    + Assign `a` to `2/3`
    + `print` `a`, and copy/paste what you see into the variable `aPrint`. Make sure it is a `character`.
    + Take a look at the documentation for `options`. Assign the value of `options()$digits` to `numDigitsStart`
    + Change the number of digits to `22`
    + Again, `print`, `a` and copy/paste what you see into the variable `aPrintv2`. Make sure it is a `character`.
    + Assign the output of `options()$digits` to `numDigitsEnd`
    
3. Floating points are weird. What gets stored might not be what you want. ["The only numbers that can be represented exactly in R’s numeric type are integers and fractions whose denominator is a power of 2."](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f) 

    + Assign the squre root of 2 to `mySqrt`
    + Print the square of this variable
    + Test (using `==`) that this variable is equal to `2`. Assign the result of this test to `isTwoRecoverable`
    + Test for near equality (using `all.equal`) that this variable is "equal" to `2`. Assign the result of this test to `closeEnough`. Make sure to read the documentation for this function because the return type can be tricky!
    
4. Let's flip some coins! Generate a thousand flips of a fair coin. Use `rbinom`, and let heads be coded as `1` and tails coded as `0`. 

    + Assign the thousand raw coin flips to a variable `flips`. Make sure the elements are integers, and make sure you flip a "fair" coin ($p=.5$).
    + Create a length `1000` `logical` `vector` called `isHeads`. Whenever you get a heads, make sure the corresponding element is `TRUE` and `FALSE` otherwise.
    + Create a variable called `numHeads` by tallying up the number of heads. 
    + Calculate the percent of time that the number changes in `flips`. Assign your number to `acceptanceRate`. Try to write only one line of code to do this.

### Easier Python Questions

1. Which Python type is ideal for each piece of data? Assign your answers to a `list` of `str`ings called `question_one`. 

    + An individual's IP address
    + whether or not an individual attended a study
    + the number of seeds found in a plant
    + the amount of time it takes for a car to race around a track

2. Floating points are weird. What gets printed is not the same as what is stored! In Python, you need to edit a class's `__str__` method if you want to control how many digits get printed, but we won't do that. Instead, we'll use [`str.format()`](https://docs.python.org/3/library/stdtypes.html#str.format) to return a string directly (instead of copy/paste-ing it). 

    + Assign `a` to `2/3`
    + `print` `a`, and copy/paste what you see into the variable `a_print`
    + Create a `str` that displays 22 digits of 2/3. Call it `a_printv2`
    + `print` the above string

3. Floating points are weird. What gets stored might not be what you want. The Python documentation has an excellent discussion of how storage behavior can be surprising. Click [here](https://docs.python.org/3/tutorial/floatingpoint.html) to read it.

    + Assign the squre root of 2 to `my_sqrt`
    + `print` the square of this variable
    + Test (using `==`) that this variable is equal to `2`. Assign the result of this test to `is_two_recoverable`
    + Test for near equality (using `np.isclose`) that this variable is "equal" to `2`. Assign the result of this test to `close_enough`. 

4. Let's flip some coins (again)! Generate a thousand flips of a fair coin. Use `np.random.binomial`, and let heads be coded as `1` and tails coded as `0`. 

    + Assign the thousand raw coin flips to a variable `flips`. Make sure the elements are integers, and make sure you flip a "fair" coin ($p=.5$).
    + Create a length `1000` `list` of `bool`s called `is_heads`. Whenever you get a heads, make sure the corresponding element is `True` and `False` otherwise.
    + Create a variable called `num_heads` by tallying up the number of heads. 
    +  Calculate the percent of time that the number changes in `flips`. Assign your number to `acceptance_rate`. Try to write only one line of code to do this.


### More Difficult R Questions


#### Examining a `vector`

  1. Say you have a vector of prices of some financial asset:

```{R, collapse = TRUE}
prices <- c(100.10, 95.98, 100.01, 99.87)
```
  
  Convert this vector into a vector of *log returns*. Call the variable `logReturns`. If $p_t$ is the price at time $t$, the log return ending at time $t$ is 
$$ r_t = \log \left( \frac{p_t}{p_{t-1}} \right) = \log p_t - \log p_{t-1}$$

  2. Do the same for *arithmetic returns*. These are regular percent changes if you scale by $100$. Call the variable `arithReturns`. The mathematical formula you need is $$ a_t = \left( \frac{p_t - p_{t-1} }{p_{t-1}} \right) \times 100 $$



#### Some Monte Carlo

Consider another **mixture density** $f(y) = \int f(y \mid x) f(x) dx$ where  

$$
Y \mid X \sim \text{Normal}(0, x^2)
$$
and

$$
X \sim \text{half-Cauchy}(0, 1).
$$
This distribution is a special case of a prior distribution that is used in Bayesian statistics [@horseshoe]. 

Suppose further that you are interested in calculating the probability that one of these random variables ends up being too far from the median: 

$$
\mathbb{P}[|Y| > 1] = \int_{y : |y| > 1} f(y)dy = \int_{y : |y| > 1} \int_{-\infty}^\infty  f(y \mid x) f(x) dx dy.
$$


  1. Simulate $X_1, \ldots, X_{5000}$ from a $\text{half-Cauchy}(0, 1)$ and call these samples `xSamps`. Hint: you can simulate from a $t$ distribution with one degree of freedom to sample from a Cauchy. Once you have regular Cauchy samples, take the absolute value of each one.
  
  2. Simulate $Y_1 \mid X_1, \ldots, Y_{5000} \mid X_{5000}$ and call the samples `ySamps`. 
  
  3. Calculate the approximate probability using `ySamps` and call it `approxProbDev1`.
  
  4. Why is simply "ignoring" `xSamps`, the samples you condition on, "equivalent" to "integrating out $x$"? Store a string response as a length $1$ character vector called `integratingOutResp`.
  
  5. Calculate another **Rao-Blackwellized** Monte Carlo estimate of $\mathbb{P}[|Y| > 1]$ from `xSamps`. Call it `approxProbDev2`. Hint: $\mathbb{P}[|Y| > 1] = \mathbb{E}[\mathbb{P}(|Y| > 1 \mid X) ]$. Calculate $\mathbb{P}(|Y| > 1 \mid X=x)$ with pencil and paper, notice it is a function in $x$, apply that function to each of `xSamps`, and average all of it together.
  
  6. Are you able to calculate an exact solution to $\mathbb{P}[|Y| > 1]$?


#### Playing Around with Some Letters


  1. Store the ordered uppercase letters of the alphabet in a length $26$ `character` `vector` called `myUpcaseLetters`. Do not hardcode this. Use a function, along with the variable `letters`.
  
  2. Create a new vector called `withReplacements` that's the same as the previous `vector`, but replace all vowels with `"---"`. Again, do not hardcode this. Find a function that searches for patterns and performs a replacement whenever that pattern is found.
  
  3. Create a length $26$ logical vector that is `TRUE` whenever an element of `letters` is a consonant, and `FALSE` everywhere else. Call it `consonant`.



### More Difficult Python Questions


### Examining a `Series` taken from a `DataFrame`

  1. Use `pd.read_csv` to correctly read in `"2013-10_Citi_Bike_trip_data_20K.csv"` as a data frame called `my_df`. Make sure to read `autograding_tips.html`.
  
  2. extract the `"starttime"` column into a separate `Series` called `s_times`
  
  3. extract date strings of those elements into a `Series` called `date_strings`
  
  4. extract time strings of those elements into a `Series` called `time_strings`

### Monte Carlo Integration

#### Approximating Probabilities

We will make use of the *Monte Carlo* [@monte-carlo-stat-methods] method below. It is a technique to approximate expectations and probabilities. If $n$ is a large number, and $X_1, \ldots, X_n$ is a random sample drawn from the distribution of interest, then 
$$
\mathbb{P}(X > 6) \approx \frac{1}{n}\sum_{i=1}^n \mathbf{1}(X_i > 6).
$$
If you haven't seen an **indicator function** before, it is defined as 

$$
\mathbf{1}(X_i > 6)
=
\begin{cases}
1 & X_i > 6 \\
0 & X_i \le 6
\end{cases}.
$$

If you wanted to visualize it, $\mathbf{1}(x > 6)$ looks like this. 


```{r, echo=FALSE, out.width="50%", fig.align = 'center'}
leX <- seq(6-5,6+5, .01)
plot(leX, leX > 6, type = "l", xlab = "x", ylab = "1(x>6)")
```

So, the sum in this expression is just a count of the number of elements that are greater than $6$. 


  1. Evaluate exactly the probability that a normal random variable with mean $5$ and standard deviation $6$ is greater than $6$. Assign it to the variable `exact_exceedance_prob` in Python.
  
  2. Simulate $1e3$ times from a standard normal distribution (mean 0 and variance 1). Call the samples `stand_norm_samps`
  
  3. Calculate a Monte Carlo estimate of $\mathbb{P}(X > 6)$ from these samples. Call it `approx_exceedance_prob1`.
  
  4. Simulate $1e3$ times from a normal distribution with mean $5$ and standard deviation $6$. Call the samples `norm_samps`. Don't use the old samples in any way.

  5. Calculate a Monte Carlo estimate of $\mathbb{P}(X > 6)$ from these new `norm_samps`. Call it `approx_exceedance_prob2`.


#### Approximating Expectations

Alternatively, we can approximate expectations. If $\mathbb{E}[f(X)]$ exists, $n$ is a large number, and $W_1, \ldots, W_n$ is a random sample drawn from the distribution of interest, then

$$
\mathbb{E}[f(W)] \approx \frac{1}{n}\sum_{i=1}^n f(W_i).
$$

Here's a new distribution. It is a **mixture distribution**, specifically a **finite mixture of normal distributions**: $f(y) =  f(y \mid X=1)P(X=1) + f(y \mid X=0)P(X=0)$ where  

$$
Y \mid X=0 \sim \text{Normal}(0, 2) \\
Y \mid X=1 \sim \text{Normal}(10, 2)
$$
and

$$
X \sim \text{Bernoulli}(.5).
$$

Both $f(y \mid X=0)$ and $f(y \mid X=1)$ are bell-curved, and $f(y)$ looks like this

```{r,  echo=FALSE, out.width="50%", fig.align = 'center'}
leX <- seq(5-20,5+20, .1)
plot(leX, .5*dnorm(leX, 0, sqrt(2)) + .5*dnorm(leX, 10, sqrt(2)), type = "l", xlab = "x", ylab = "f(y)")
```


  1. Evaluate exactly $\mathbb{E}[Y]$. Assign it to the variable `exact_mean` in Python.
  
  2. Simulate $1e3$ times from the Bernoulli distribution. Call the samples `bernoulli_flips`

  3. Simulate $Y_1 \mid X_1, \ldots, Y_{1000} \mid X_{1000}$ and call the samples `cond_norm_samps`. 
  
  4. Calculate a Monte Carlo estimate of $\mathbb{E}[Y]$ from `cond_norm_samps`. Call it `approx_ave_1`. Why is simply "ignoring" `bernoulli_flips`, the samples you condition on, "equivalent" to "integrating them out?" 
  
  5. Calculate a **Rao-Blackwellized** Monte Carlo estimate of $\mathbb{E}[Y]$ from `bernoulli_flips`. Call it `approx_ave_2`. Hint: $\mathbb{E}[Y] = \mathbb{E}[\mathbb{E}(Y \mid X) ]$. Calculate $\mathbb{E}(Y \mid X_i)$ exactly, and evaluate that function on each $X_i$ sample, and then average them together. Rao-Blackwellization is a variance-reduction technique that allows you come up with lower-variance estimates given a fixed computational budget.
  
  

<!-- ### Exercises -->

<!-- #### Easy -->

<!-- All answers to questions related to R should be written in a file named `easy_data_types_exercises.R`. All answers to questions related to Python should be written in a file named `easy_data_types_exercises.py`. -->


<!-- 1. Which Python type is ideal for each piece of data? Assign your answers to a `list` of `str`ings called `question_one`.  -->

<!--     + An individual's IP address -->
<!--     + whether or not an individual attended a study -->
<!--     + the number of seeds found in a plant -->
<!--     + the amount of time it takes for a car to race around a track -->

<!-- 2. Answer the same question above, but use R types? Assign your answers to a `character vector` of length four called `questionOne`.  -->

<!--     + An individual's IP address -->
<!--     + whether or not an individual attended a study -->
<!--     + the number of seeds found in a plant -->
<!--     + the amount of time it takes for a car to race around a track -->

<!-- 3. ["The only numbers that can be represented exactly in R’s numeric type are integers and fractions whose denominator is a power of 2."](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Why-doesn_0027t-R-think-these-numbers-are-equal_003f) Provide ten examples of numbers that have nonzero decimal components and are not exactly equal represented by R's `numeric` type. Store them all in a vector called `notExactFloats`. -->

