# Functional Programming

## FP Background

Even though this book only discusses one of our languages, the motivation for **functional programming** in both R and Python is clearly in [Advanced R](https://adv-r.hadley.nz/fp.html):

> A functional style tends to create functions that can easily be analysed in isolation (i.e. using only local information), and hence is often much easier to automatically optimise or parallelise.

This sounds like a good thing to aspire to, so what is functional programming? Unfortunately, it's hard to give a one-line definition that is also comprehensive. Even with a definition, it isn't immediately obvious why the pieces of that definition would be conducive to the above goals. 

Here is my definition: 

- a functional programming style is a style of programming that favors using functions that are both *first-class* and *pure*. 

Neither R nor Python is a 100% pure functional language. So for us, it's a style that we can choose to let us guide us, or that we can disregard. You can choose to employ a more functional style, or you can choose to use elements discussed in \@ref(an-introduction-to-object-oriented-programming). Some people unilaterally prefer one to the other, but others prefer to decide which to use depending on the task at hand. 

1. **First class functions** are functions that can be passed as arguments to other functions, can be returned from other functions, and can be assigned to variables or stored in data structures. TODO quote. 

Functions in both R and Python are first class. Many of the examples we discuss in this chapter would not exist if they were not. 

2. **Pure functions** return the same output if they are given the same input, and they do not produce **side-effects**. 

Side-effects are changes made to non-temporary variables, to the "state" of the program. If your function refers to or modifies global variables, or if it modifies arguments passed in, then it is more difficult to understand. As was discussed in TODO, this is possible in both R and Python, but the function is not completely understandable 

|   a.) by itself in isolation, or 

|   b.) without knowing how many times it is called. 

Regarding a.), we discussed examples where changing the value of a single global variable can completely change the behavior of an entire program (not just a single function!). Regarding b.) a variable has **referential transparency** if it can be replaced with a single value without changing the program's behavior. 

3. **Recursion versus Iteration**

4. **Lazy versus Strict Evaluation**



```{block, type='rmd-details'}
imperative versus procedural
```



## Apply Functions in R

There are several functions in R whose names end in "apply". The ones we discuss are `sapply`, `vapply`, `lapply`, `apply`, `tapply` and `mapply`. Each of these takes a function as one of its arguments. This is possible because R has first-class functions.


```{r, echo=F}
myDFrows <- 10
myDFcols <- 100
myDF <- data.frame(matrix(rnorm(myDFrows*myDFcols), ncol = myDFcols))
```

Suppose we have a `data.frame` that has `r nrow(myDF)` rows and `r ncol(myDF)` columns. What if we want to take the mean of each column?

An amateurish way to do this would be something like the following.

```{r, eval=F}
myFirstMean <- mean(myDF[,1])
mySecondMean <- mean(myDF[,2])
myThirdMean <- mean(myDF[,3])
# so on and so forth
```

You need one line of code for each column in this data frame. For data frames with a lot of columns, this becomes quite tedious. You should also ask yourself what happens to you and your collaborators when the data frame changes even slightly, or if you want to apply a different function to its columns. Third, the results are not stored in a single container. You are making it difficult on yourself if you want to use these variables in subsequent pieces of code.  TODO cite DRY

Instead, prefer the use of `sapply` in this situation. The "s" in `sapply` stands for "simplified." In this bit of code `mean` is called on each column of th data frame. `sapply` applies the function over columns, instead of rows, because data frames are internally a `list` of columns. 

```{r, collapse=TRUE}
myMeans <- sapply(myDF, mean)
head(myMeans)
```

Each call to `mean` returns a `double` `vector` of length $1$. This is necessary if you want to collect all the results into a `vector`--remember, all elements of a `vector` have to have the same type. To get the same behavior, you might also consider using `vapply(myDF, mean, numeric(1))`.

In the above case, "simplify" referred to how one-hundred length-$1$ vectors were simplified into one length-$100$ vector. However, "simplified" does not necessarily imply that all elements will be stored in a vector. Consider the `summary` function, which returns a `double` `vector` of length $6$. In this case, one-hundred length-$6$ vectors were simplified into one $6 \times 100$ matrix.

```{r, collapse = TRUE}
mySummaries <- sapply(myDF, summary)
is.matrix(mySummaries)
dim(mySummaries)
```


For functions that don't return amenable types that fit into a `vector`, `matrix` or `array`, they might need to be stored in `list`. In this situation, you would need `lapply`. The "l" in `lapply` stands for "list". `lapply` always returns a `list` of the same length as the `input`.

```{r, collapse = TRUE}
regress <- function(y){ lm(y ~ 1) }
myRegs <- lapply(myDF, regress)
length(myRegs)
class(myRegs[[1]])
summary(myRegs[[12]])
```

I use `sapply` and `lapply` the most, personally. The next most common function I use is `apply`. I use it to apply functions to *rows* instead of columns, although, it can also apply functions over columns, just as the other functions we discussed can. 

```{r, collapse = TRUE}
dim(myDF)
apply(myDF, 1, mean)
```

`tapply` can be very handy when you need it. First, we've alluded to the definition before in TODO, but a **ragged array** is a collection of arrays that all have potentially different lengths. I don't typically construct such an object and then pass it to `tapply`. Rather, I let `tapply` construct the ragged array for me. The first argument it expects is "typically vector-like", while the second tells us how to break that `vector` into chunks. The third argument is a function that gets applied to each `vector` chunk.

If I wanted the average home price for each city, I could use something like this.
```{r, collapse = TRUE}
albRealEstate <- read.csv("data/albemarle_real_estate.csv")
head(albRealEstate)
unique(albRealEstate$City)
tapply(albRealEstate$TotalValue, list(albRealEstate$City), mean)
```
You might be wondering why we put `albRealEstate$City` into a `list`. That seems kind of unnecessary. This is because `tapply` can be used with multiple `factor`s--this will break down the `vector` input into a finer partition. The second argument must be one object, though, so all of these `factor`s must be collected into a `list`. The following code produces a "pivot table."

```{r, collapse = TRUE}
tapply(albRealEstate$TotalValue, list(albRealEstate$City, albRealEstate$Condition), mean)
```

For situations where TODO, you also have `by` and `aggregate`.

The [documentation of `mapply`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/mapply.html) states `mapply` is a multivariate version of `sapply`. `sapply` worked with univariate functions: the function was called multiple times, but each time with a single argument. If you have a function that takes multiple arguments, and you want those arguments to change each time the function is called, then you might be able to use `mapply`. 

Here is a short example. Regarding the `n=` argument of `rnorm`, the documentation explains, "[i]f `length(n) > 1`, the length is taken to be the number required." This would be a problem if we want to sample a.) three times from a mean $0$ normal, b.) twice from a mean $100$ normal, and c.) once from a mean $-100$ normal distribution.

```{r}
rnorm(n = c(3,2,1), mean = c(0,100,-100), sd = c(.01, .01, .01))
mapply(rnorm, n = c(3,2,1), mean = c(0,100,-100), sd = c(.01, .01, .01))
```


Consider another less contrived example. Suppose we want to plot a bivariate Gaussian distribution. 

$$
f(x,y) = \frac{1}{2 \pi}  \exp\left[ -\frac{x ^2 + y^2}{2}     \right]
$$
The random elements $x$ and $y$, in this particular case, are uncorrelated, each have unit variance, and zero mean. This density is a surface in 3-d dimensional space. To visualize this, we would need to

1. generate a "grid" of points in $\mathbb{R}^2$, 
2. evaluate our function on each point, and then
3. call some plotting function that takes this all and makes a pretty picture.

There are a couple ways you could write this function. One way might take two arguments, and another might take one argument. If we are to use `mapply`, we need the function to take two arguments.

```{r, collapse = TRUE}
fTwoArgs <- function(x,y){
  exp(-.5*(x^2 + y^2)) / 2 / pi
}
```

We can construct every possible point on a grid with the `expand.grid` function.

```{r, collapse = TRUE}
xGrid <- seq(-3,3,.1)
yGrid <- seq(-3,3,.1)
grid <- expand.grid(xGrid, yGrid)
head(grid)
```

`mapply` would take `fTwoArgs`, and effectively call it on every row pair. The pairs do not need to be organized in a `data.frame`, though.

```{r, collapse=TRUE}
funcOut1 <- mapply(fTwoArgs, grid[,1], grid[,2])
head(funcOut1)
rectangularOutput <- matrix(funcOut1, ncol = length(xGrid))
persp(xGrid, yGrid, rectangularOutput, 
      zlab = "f(x,y)", xlab = "x", ylab = "y")
```

If you prefer using `apply`, that is also possible, but you would need to rewrite the function to take one (length-two) argument. 

```{r, collapse=T}
fOneArg <- function(vec){
  exp(-sum(vec^2)/2)/2/pi
} 
funcOut2 <- apply(grid, 1, fOneArg)
moreRectOut <- matrix(funcOut2, ncol = length(xGrid))
contour(xGrid, yGrid, moreRectOut, xlab = "x", ylab = "y")
```



## Selected Apply Functions in Python


Numpy provides a [number of functions](https://numpy.org/doc/stable/reference/routines.functional.html) that facilitate working with `np.ndarray`s in a functional style. For example, [`np.apply_along_axis`](https://numpy.org/doc/stable/reference/generated/numpy.apply_along_axis.html) is similar to R's `apply`. `apply` had a `MARGIN` argument (`1` sums rows, `2` sums columns), whereas this function has a `axis=` argument (`0` sums columns, `1` sums rows). 

```{python, collapse = TRUE}
import numpy as np
my_array = np.arange(6).reshape((2,3))
my_array
np.apply_along_axis(sum, 0, my_array) # summing columns
np.apply_along_axis(sum, 1, my_array) # summing rows
```



```{python, collapse = TRUE}
my_array = np.random.normal(size=(10,1000))
np.apply_along_axis(sum, 0, my_array).shape 
np.apply_along_axis(sum, 1, my_array).shape
```

Pandas `DataFrame`s have an [`.apply` method](https://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.apply.html) that is very similar to `apply` in R,^[You should know that a lot of special-case functions that you typically apply to rows or columns come built-in as `DataFrame` methods. For instance, `.mean` would allow you to do something like `my_df.mean()`.] but again, just like the above function, you have to think about an `axis=` argument instead of a `MARGIN=` argument. 

```{python, collapse = TRUE}
import pandas as pd
alb_real_est = pd.read_csv("data/albemarle_real_estate.csv")
alb_real_est.shape
alb_real_est.apply(len, axis=0) # length of columns
type(alb_real_est.apply(len, axis=1)) # length of rows
```

Another thing to keep in mind is that `DataFrame`s, unlike `ndarray`s, don't have to have the same type for all elements. If you have mixed column types, then summing rows, for instance, might not make sense. This just requires subsetting columns before `.apply`ing a function to rows. Here is an example of computing each property's "score". 

```{python, collapse = TRUE}
import pandas as pd
# alb_real_est.apply(sum, axis=1) # can't add letters to numbers!
def get_prop_score(row):
  return 2*row[0] + 3*row[1]
alb_real_est['Score'] = alb_real_est[['FinSqFt','LotSize']].apply(get_prop_score, 1)
alb_real_est[['FinSqFt','LotSize','Score']].head(2)
```

`.apply` also works with more than one function at a time.

```{python, collapse = TRUE}
alb_real_est[['FinSqFt','LotSize']].apply([sum, len])
```

If you do not want to waste two lines defining a function with `def`, you can use an anonymous (unnamed) [**lambda function**](https://docs.python.org/3/tutorial/controlflow.html#lambda-expressions). Be careful, though--if your function is complex enough, then your lines will get quite wide. For instance, this example is pushing it.

```{python, collapse = TRUE}
alb_real_est[['FinSqFt','LotSize']].apply(lambda row : sum(row*[2,3]), 1)[:4]
```


If you want to apply a (scalar-valued) function that takes only individual elements, you should try to use a unary function (recall that this was discussed in TODO). If no such unary function exists, you can apply it with [`.applymap`](https://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.applymap.html#pandas.DataFrame.applymap).

```{python, collapse = TRUE}
alb_real_est[['FinSqFt','LotSize']].applymap(lambda e : e + 1).head(2)
```


Last, we have a [`.groupby`](https://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.groupby.html) method, which can be used to mirror the behavior of R's `tapply`, `aggregate` or `by`. It can take the `DataFrame` it belongs to, and group its rows into multiple sub-`DataFrame`s. The collection of sub-`DataFrames` has a lot of the same methods that an individual `DataFrame` has (e.g. the subsetting operators,  and the `.apply` method), which can all be used in a second step of calculating things on each sub-`DataFrame`.

```{python, collapse = TRUE}
type(alb_real_est.groupby(['City']))
type(alb_real_est.groupby(['City'])['TotalValue'])
```

Here is an example that models some pretty typical functionality. It shows two ways to get the average home price by city. The first line groups the rows by which `City` they are in, extracts the `TotalValue` column in each sub-`DataFrame`, and then `.apply`s the `np.average` function on the sole column found in each sub-`DataFrame`. The second applys a lambda function to each sub-`DataFrame` directly. 

```{python, collapse = TRUE}
alb_real_est.groupby(['City'])['TotalValue'].apply(np.average)
alb_real_est.groupby(['City']).apply(lambda df : np.average(df['TotalValue']))
```

More tips on this programming pattern can be found [here](https://pandas.pydata.org/pandas-docs/stable/user_guide/groupby.html).


## Reduce

Both R and Python have reduce functions. R has `Reduce`, while Python has `reduce`.




### Exercises

.apply with weird predicate function

Optimization example 

Importance sampling example

