# Data Frames

The rectangular array (e.g. an Excel spreadsheet ) of information is what many think of when they hear the word "data." Each column contains elements of a shared data type, and these data types can vary from column to column. 

There is a type for this in R and Python: a data frame. It might even be the most common way that data is stored in both R and Python programs because many functions that read in data from an external source return objects of this type (e.g. `read.csv()` in R and `pd.read_csv()` in Python). 

R and Python's data frames have a number of things in common:

  1. each column must be the same length as all other columns,
  2. each column's elements will all have the same type, 
  3. elements in any rows can have different types, 
  4. columns and rows can be named in different ways, 
  5. there are many ways to get and set different subsets of data, and
  6. when reading in data, the same sorts of difficulties arise in both languages.


## Data Frames in R


Let's consider as an example Fisher's "Iris" data set [@misc_iris_53]. We will read this data set in from a comma separated file (more information on input/output can be found in chapter \@ref(input-and-output)). This file can be downloaded from this link: [https://archive.ics.uci.edu/ml/datasets/iris](https://archive.ics.uci.edu/ml/datasets/iris). 


```{r, collapse=TRUE}
irisData <- read.csv("data/iris.csv", header = F)
head(irisData, 3)
typeof(irisData)
class(irisData) # we'll talk more about classes later
dim(irisData)
nrow(irisData)
ncol(irisData)
```

There are some exceptions, but most data sets can be stored as a `data.frame`. These kinds of two-dimensional data sets are quite common. Any particular row is often an observation on one experimental unit (e.g. person, place or thing). Looking at a particular column gives you one kind of measurement stored for all observations. 

::: {.rmd-caution}
Do not rely on the default arguments of `read.csv()` or `read.table()`! After you read in a data frame, always check to make sure that 

1. the number of columns is correct because the correct column *separator* was used (c.f. `sep=`),
2. column names were parsed correctly, if there were some in the raw text file,
3. the first row of data wasn't used as a column name sequence, if there weren't column names in the text file, and
4. the last few rows aren't reading in empty spaces
5. character columns are read in correctly (c.f. `stringsAsFactors=`), and
6. special characters signifying missing data were correctly identified (c.f. `na.strings=`).
:::


[A `data.frame` is a special case of a `list`](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Data-frame-objects). Every element of the list is a column. Columns can be `vector`s or `factor`s, and they can all be of a different type. This is one of the biggest differences between data frames and `matrix`s. They are both two-dimensional, but a `matrix` needs elements to be all the same type. Unlike a general `list`, a `data.frame` requires all of its columns to have the same number of elements. In other words, the `data.frame` is not a "*ragged*" list.


Often times you will need to extract pieces of information from a `data.frame`. This can be done in many ways. If the columns have names, you can use the `$` operator to access a single column. Accessing a single column might be followed up by creating a new vector. You can also use the `[` operator to access multiple columns by name.  

```{r, collapse=TRUE}
colnames(irisData) <- c("sepal.length", "sepal.width", 
                        "petal.length","petal.width", 
                        "species")
firstCol <- irisData$sepal.length
head(firstCol)
firstTwoCols <- irisData[c("sepal.length", "sepal.width")] 
head(firstTwoCols, 3)
```

The `[` operator is also useful for selecting rows and columns by index numbers, or by some logical criteria.

```{r, collapse=TRUE}
topLeft <- irisData[1,1] # first row, first col
topLeft
firstThreeRows <- irisData[1:3,] # rows 1-3, all cols
firstThreeRows
setosaOnly <- irisData[irisData$species == "Iris-setosa",] # rows where species column is setosa
head(setosaOnly, 3)
```

In the code above, `irisData$species == "Iris-setosa"` creates a logical vector (try it!) using the vectorized `==` operator. The `[` operator selects the rows for which the corresponding element of this logical vector is `TRUE`. 

::: {.rmd-details}
Be careful: depending on how you use the square brackets, you can either get a `data.frame` or a `vector.` As an example, try both `class(irisData[,1])` and `class(irisData[,c(1,2)])`. 
:::


In R, `data.frame`s might have row names. You can get and set this character `vector` with the `rownames()` function. You can access rows by name using the square bracket operator. 

```{r, collapse = TRUE}
head(rownames(irisData))
rownames(irisData) <- as.numeric(rownames(irisData)) + 1000
head(rownames(irisData))
irisData["1002",]
```


Code that modifies data usually looks quite similar to code extracting data. You'll notice a lot of the same symbols (e.g. `$`, `[`, etc.), but the (`<-`) will point in the other direction. 


```{r, collapse = TRUE}
irisData$columnOfOnes <- rep(1, nrow(irisData))
irisData[1:2,1] <- rnorm(n = 2, mean = 999)
irisData[,'sepal.width'] <- rnorm(n = nrow(irisData), mean = -999)
irisData[irisData$species == "Iris-setosa", 'species'] <- "SETOSA!"
head(irisData,3)
```


## Data Frames in Python


The Pandas library in Python has data frames that are modeled after R's. 

```{python, collapse=TRUE}
import pandas as pd
iris_data = pd.read_csv("data/iris.csv", header = None)
iris_data.head(3)
iris_data.shape
len(iris_data) # num rows
len(iris_data.columns) # num columns
list(iris_data.dtypes)
```

The structure is very similar to that of R's data frame. It's two dimensional, and you can [access columns and rows by name or number.](https://pandas.pydata.org/pandas-docs/stable/user_guide/indexing.html) Each column is a `Series` object, and each column can have a different `dtype`, which is analogous to R's situation. Again, because the elements need to be the same type along columns only, this is a big difference between 2-d Numpy `ndarray`s and `DataFrame`s (c.f. R's `matrix` versus R's `data.frame`). 



::: {.rmd-caution}
Again, do not rely on the default arguments of `pd.read_csv()`! After you read in a data set, always check that

1. the number of columns is correct because the correct column *separator* was used (c.f. `sep=`),
2. column names were parsed correctly, if there were some in the raw text file,
3. the first row of data wasn't used as a column name sequence, if there weren't column names in the text file (c.f. `header=`), and
4. the last few rows aren't reading in empty spaces
5. character columns are read in correctly (c.f. `dtype=`), and
6. special characters signifying missing data were correctly identified (c.f. `na.values=`).
:::


Square brackets are a little different in Python than they are in R. Just like in R, you can access columns by name with square brackets, and you can also access rows. Unlike R, though, you don't have to specify both rows and columns every time you use the square brackets.

```{python, collapse=TRUE}
iris_data.columns = ["sepal.length", "sepal.width", 
                     "petal.length", "petal.width", 
                     "species"]
first_col = iris_data['sepal.length']
first_col.head()
first_two_cols = iris_data[["sepal.length", "sepal.width"]]
first_two_cols.head(3)
iris_data[0:2]
```


::: {.rmd-details}
Notice that `iris_data['sepal.length']` returns a `Series` and `iris_data[["sepal.length", "sepal.width"]` returns a Pandas `DataFrame`. This behavior is similar to what happened in R's. For more details, click [here](https://pandas.pydata.org/pandas-docs/stable/user_guide/dsintro.html#indexing-selection).
::: 

You can select columns and rows by number with the [`.iloc` method](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.iloc.html). `iloc` is (probably) short for "integer location." 

```{python, collapse=TRUE}
# specify rows/cols by number
top_left = iris_data.iloc[0,0]
top_left
first_three_rows = iris_data.iloc[:3,]
first_three_rows
#setosa_only = iris_data[irisData$species == "Iris-setosa",]  # easieriwith loc?
#head(setosaOnly)
```

Selecting columns by anything besides integer number can be done with the [`.loc()` method](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.loc.html). You should generally prefer this method to access columns because accessing things by *name* instead of *number* is more readable. Here are some examples. 

```{python, collapse=TRUE}
sepal_w_to_pedal_w = iris_data.loc['sepal.width':'pedal.width']
sepal_w_to_pedal_w
setosa_only = iris_data.loc[iris_data['species'] == "Iris-setosa",]
setosa_only.head(3)
```

Notice we used a `slice` (i.e. `'sepal.width':'pedal.width'`) to access many columns by only referring to the left-most and the right-most. Unlike slicing with numbers, [the right end is included](https://pandas.pydata.org/pandas-docs/stable/user_guide/indexing.html#slicing-ranges). Also note that this does not work with the regular square bracket operator (i.e. `iris_data['sepal.width':'pedal.width']`). The second example filters out the rows where the `"species"` column elements are equal to `"Iris-setosa"`.

Each `DataFrame` in pandas comes with an `.index` attribute. This is analogous to a row name in R, but it's much more flexible because the index can take on a variety of types. This can help us highlight the difference between `.loc` and `.iloc`. Recall that `.loc` was *label-based selection.* Labels don't necessarily have to be strings. Consider the following example.

```{python, collapse = TRUE}
iris_data.index
iris_data = iris_data.set_index(iris_data.index[::-1]) # reverse the index
iris_data.tail(2) # top is now bottom
iris_data.loc[0] # last row has 0 index
iris_data.iloc[0] # first row with big index 
```

`iris_data.loc[0]` selects the `0`th index. The second line reversed the indexes, so this is actually the last row. If you want the first row, use `iris_data.iloc[0]`.

<!-- [`DataFrame`s have a huge amount  many useful attributes and methods, too.](https://pandas.pydata.org/docs/reference/api/pandas.DataFrame.html#) -->

Modifying data inside a data frame looks quite similar to extracting data. You'll recognize a lot of the methods mentioned earlier. 

```{python, collapse = TRUE}
import numpy as np
n_rows = iris_data.shape[0]
iris_data['col_ones'] = np.repeat(1.0, n_rows)
iris_data.iloc[:2,0] =  np.random.normal(loc=999, size=2)
iris_data.loc[:,'sepal.width'] = np.random.normal(loc=-999, size=n_rows)
iris_data.loc[iris_data['species'] == "Iris-setosa", 'species'] = "SETOSA!"
iris_data.head(3)
```


TODO example of assigning a Series with different index.
