# Working With Text Data

We have already talked a little about how to work with text data in this book. Regarding Python, section \@ref(vectorization-in-python) mentioned that Pandas `Series` objects have a [`.str` accessor attribute](https://pandas.pydata.org/pandas-docs/version/1.3/user_guide/text.html#string-methods) that has plenty of special methods that will work on string data. The same tools can be used whether or not these `Series` objects are contained in a Pandas `DataFrame`. 

Regarding R, `character` `vector`s were mentioned in chapter \@ref(r-vectors-versus-numpy-arrays-and-pandas-series). There are many functions that operate on these, too, whether or not they are held inside an R `data.frame`. They might be a little harder to find because they aren't methods, so pressing `<Tab>` and using your GUI's autocomplete feature doesn't reveal them as easily.

Suppose you're interested in replacing lowercase letters with uppercase ones, removing certain characters from text, or counting the number of times a certain expression appears. Up until now, as long as you can find a function or method that performs the task, you were doing just fine. If you need to do something with text data, there's probably a function for it. 

Notice what all of these tasks have in common--they all require the ability to find patterns. When your patterns are easy to describe (e.g. find all lowercase "a"s), then all is well. What can make matters more complicated, however, is when the patterns are more difficult to describe (e.g. find all valid email addresses). That is why this chapter is primarily concerned with discussing **regular expressions,** which are a tool that help you describe the patterns in text.


## An Introducton to Regular Expressions

### Literal Characters versus Metacharacters

Every character in a regular expression is interpreted in one of two ways. Either it is interpreted as a 

  1. literal character, or as a 
  2. metacharacter.
  
If it is a literal character, then the character is the *literal* pattern. For example, in the regular expression "e", the character "e" has a literal interpretation. If you seek to capitalize all instances of "e" in the following phrase, you can do it pretty easily. As long as you know which function performs find-and-replace, you're good. The pattern is trivial to specify. 

On the other hand, if I asked you to remove `$`s from price or salary data, you might have a little more difficulty. This is because `$` is a *metacharacter* in regular expressions, and so it has a special meaning.^[The dollar sign is useful if you only want to find certain patterns that finish a line. It takes the characters preceding it, and says, only look for that pattern if it comes at the end of a string.] In the examples below, if `$` is interpreted as a regular expression, the pattern will not be found at all, despite the prevalence of *literal* dollar signs. 

There are a few functions in R that perform find-and-replace, but in this case, I use `gsub()`. In Pandas, I can use [`.str.replace()`](https://pandas.pydata.org/docs/reference/api/pandas.Series.str.replace.html), to do this. Here are the examples that find patterns that are described by literal characters. 

```{r, collapse = TRUE}
# in R
gsub(pattern = "e", replacement = "E", x = "I don't need a regex for this!")
```

```{python, collapse = TRUE}
# in Python
import pandas as pd
pd.Series(["I don't need a regex for this!"]).str.replace(pat="e",repl="E")
```

On the other hand, here are a few examples that remove dollar signs. We generally have two options to recognize symbols that happen to be metacharacters. 

  1. We can *escape* the dollar sign. That means you need to put a backslash (i.e. `\`) before the dollar sign. The backslash is a metacharacter looks at the character coming after it, and it either removes the special meaning from a metacharacter, or adds special meaning to a literal character. 
  2. Alternatively, we can tell the function to ignore regular expressions. `gsub()` can take `fixed=TRUE`, and `.str.replace()` can take `regex=False`. 
  

```{python, collapse = TRUE}
# in Python
pd.Series(["$100, $200"]).str.replace(pat="$",repl="",regex=False)
pd.Series(["$100, $200"]).str.replace(pat="\$",repl="")
```

```{r, collapse = TRUE}
# in R
gsub(pattern = "$", replacement = "", x = c("$100, $200"), fixed=TRUE)
stringr::str_remove_all(c("$100, $200"), pattern = "\\$")
```

  
### The Trouble With Backslashes: Escape Sequences

You might have noticed above and gotten confused--sometimes in Python and in R, we need *two* backslashes instead of *one*. This is because backslashes have another purpose that can complicate our using them to escape metacharacters. They also help us write untypeable characters, also known as **escape sequences**. We need to be able to do this even if we aren't using regular expressions. 

For instance, consider the way we type the "newline" character. Even though it is understood by the computer as one character, it takes us two keystrokes to write it with our keyboard. `\` is one character, and `n` is another, but together they are one!

```{r, collapse = TRUE}
nchar('\n') #in R
```

```{python, collapse = TRUE}
len('\n') #in Python
```

`str`s in Python and `character` `vectors` in R will look for these combinations by default. When we specify regular expressions with strings, the backslashes will be used first for this purpose. Their regular expression purpose is a second priority. 

The reason we used `\\$` in the above example is to escape the second backslash. `\$` is not a special character, but Python and R will handle it differently. Python will not recognize it, and it won't complain that it didn't. On the other hand, R will throw an error that it can't recognize it. 

```{python, collapse = TRUE}
len('\$') # in Python, not special
```

```{r, error = TRUE, collapse = TRUE}
nchar('\$') # R gets confused
```

There is another way to deal with this issue--**raw strings!** Raw strings make life easier because they do not interpret backslashes as the beginning of escape sequences. You can make them in R and Python by putting an "r" in front of the quotation marks. However, it is slightly more complicated in R because you need a delimiter pair inside the quotation marks--for more information type `?Quotes` in your R console.

```{python, collapse = TRUE}
len(r'\n') # in Python 
```

```{r, error = TRUE, collapse = TRUE}
nchar(r'{\$}') # in R
```


### More Examples of Using Regular Expressions




TODO
  - raw strings
  - text in files or finding files by name
  - backslashes are a nightmare
  - citations

