--- 
title: "An Introduction to `R` and `Python` For Data Analysis: a Side By Side Approach"
author: "Taylor R. Brown"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
#output: bookdown::gitbook
bookdown::pdf_book:
    citation_package: biblatex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: tbrown122387/r_and_python_book
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Preface

## Who this book is for

This book is written for a graduate level class that introduces students to both `R` and `Python` for analyzing and modeling data. This text does not describe statistical modeling techniques in any detail (although many exercises will be motivated by different statistical techniques), but rather teaches the basics of data "cleaning" and "munging." After all, it is this aspect that is common to all data analysis tasks, and it is usually the the first step in the process.

## Why the side by side approach?

The `R` versus `Python` holy war notwithstanding, I have been a part of several institutions that desire to teach their students *both* `R` and `Python`. I think teaching them side by side like this helps reinforce shared concepts, and is easier on every student's memory. 

## Conventions

Sometimes `R` and `Python` code look very similar. Sometimes, they even look identical. This is why we usually separate `R` and `Python` code into separate sections. 

However, whenever it is necessary to prevent confusion, I will remind you what language is being used in comments (more about comments in section TODO). 

```{python}
# in python
print('hello world')
```


```{r}
# in R
print('hello world')
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


## Installing the Required Software

To get started, you must install both `R` and `Python`. The installation process depends on what type of operating system your machine is running.

There are many options available to you for how to do this. These options are likely to change over time, and different options may be preferable to different people. I provide one recommendation below. If you prefer to take a different route, or you already have a solution to either of these, feel free to keep using it. The code should still work as long as you're using relatively recent versions of each.

### Installing `R` (and `RStudio`)

It is recommended that you install `R` *and* `RStudio Desktop`. `RStudio Desktop` is a graphical user interface with many tools that making writing `R` easier and more fun.

Install `R` from the Comprehensive R Archive Network (CRAN). You can access instructions for your specific machine by clicking [here.](https://cran.r-project.org/) 

You can get `RStudio Desktop` directly from the [company's website](https://www.rstudio.com/products/rstudio/download/#download).

### Installing `Python` by Installing `Anaconda`

It is recommended that you install `Anaconda`, which is a package manager, environment manager, and Python distribution with many third party open source packages. It provides a graphical user interface for us, too, just as `RStudio` does. You can access instructions for your specific machine and OS by clicking [here](https://docs.anaconda.com/anaconda/install/#).


## Automatic Grading

If you are an instructor, you may contact TODO to receive files that allow you to implement programmatic grading for this course. `R` scripts are graded with the `gradeR` package, and `Python` scripts with the `Otter Grader` package (TODO cite).





