--- 
title: "An Introduction to R and Python For Data Analysis: a Side By Side Approach"
author: "Taylor R. Brown"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
#output: bookdown::gitbook
bookdown::pdf_book:
    citation_package: biblatex
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: tbrown122387/r_and_python_book
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# Preface

## About this book

The `R` versus `Python` holy war notwithstanding, many organizations and institutions aim to teach their graduate-level students *both* `R` and `Python`. This book serves this purpose. 

If you are using them for analyzing data, `R` and `Python` do a lot of the same stuff in pretty similar ways, so it does not always make sense to teach one language after the other. Imagine learning about vectorization in `R`, and then several weeks later learning about the same concept in `Python`. You might end up spending a lot of time refreshing your memory before getting started for the second time. The side-by-side approach ought to help reinforce shared concepts. In my opinion, it also helps to highlights key differences.

This text does not describe statistical modeling techniques in detail, although many exercises will be motivated by different statistical techniques. Rather, it teaches the basics of data "cleaning", "munging" and manipulation. 

This book is written for aspiring data scientists, not necessarily computer scientists. Why do I draw the distinction? When discussing different types, for example, I do not discuss data structures in any depth. Rather, I discuss examples of applications where different types would be most useful. 

It is meant to be read on a computer with an internet connection. There are plenty of hyperlinks to click. 

This book does not attempt to be an authoritative reference. This is my attempt at balancing depth and breadth for a one-semester course. Plenty of discovery will be left to the reader. 

Also, I imagine it being used as a reference. It's more important that students do the exercises on their own. Associating exercises with a (sub-)section in the textbook will give them support when they need it.

Chapters and sections are meant to be read in order, too. Say, for instance, a topic in `R` is discussed first. If the `Python` discussion comes second, that discussion will reference and make comparisons with details mentioned in the `R` section. So read sections in order. 

I believe it is more effective if students read this text on their own, instead of me giving lectures on the material. Personally, I find it difficult to give lectures that both a.) have a lot of detail on programming, and b.) are still captivating. It's also a "teach a man how to fish" kind of thing. After this course ends, the student should be trained in being self-sufficient, otherwise he/she will have a hard time in "real life" programming. TODO edit this.

## Conventions

Sometimes `R` and `Python` code look very similar. Sometimes, they even look identical. This is why we usually separate `R` and `Python` code into separate sections. 

However, whenever it is necessary to prevent confusion, I will remind you what language is being used in comments (more about comments in \@ref(hello-world-in-python) ). 

```{python, collapse = TRUE}
# in python
print('hello world')
```


```{r, collapse = TRUE}
# in R
print('hello world')
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


## Installing the Required Software

To get started, you must install both `R` and `Python`. The installation process depends on what kind of machine you have (e.g. what type of operating system your machine is running, is your processor $32$ or $64$ bit, etc.).

I provide some helpful links below. However, there are many alternatives available to you for how to install, and different options may be preferable to different students. Moreover, these options are likely to change over time. 

If you prefer to take a different route, or you already have a solution to either of these, feel free to keep using it. The code should still work as long as you're using relatively recent versions of each. For Python, you need a version that is `>=3.6`, and for `R`, it nees to be `>=4.0.0`.

### Installing `R` (and `RStudio`)

It is recommended that you install `R` *and* `RStudio Desktop`. `RStudio Desktop` is a graphical user interface with many tools that making writing `R` easier and more fun.

Install `R` from the Comprehensive R Archive Network (CRAN). You can access instructions for your specific machine by clicking [here.](https://cran.r-project.org/) 

You can get `RStudio Desktop` directly from the [company's website](https://www.rstudio.com/products/rstudio/download/#download).

### Installing `Python` by Installing `Anaconda`

It is recommended that you install `Anaconda`, which is a package manager, environment manager, and Python distribution with many third party open source packages. It provides a graphical user interface for us, too, just as `RStudio` does. You can access instructions for your specific machine and OS by clicking [here](https://docs.anaconda.com/anaconda/install/#).


## Automatic Grading

If you are an instructor, you may contact TODO to receive files that allow you to implement programmatic grading for this course. `R` scripts are graded with the `gradeR` package, and `Python` scripts with the `Otter Grader` package (TODO cite).





