# Preface {-}

## About this book {-}

This book is written to be used in a one-semester statistical computing class that teaches both R and Python to graduate students in a statistics or data science department. This book is written for students that do not necessarily possess any previous familiarity with writing code. 

 * If you are using them for analyzing data, R and Python do a lot of the same stuff in pretty similar ways, so it does not always make sense to teach one language after the other. Imagine learning about vectorization, say, in R, and then several weeks later learning about the same concept in Python. You might end up spending a lot of time refreshing your memory before getting started for the second time. The side-by-side approach ought to help reinforce shared concepts. In my opinion, it also helps to highlight key differences.

 * This text does not describe statistical modeling techniques in detail, although many exercises will be motivated by different statistical techniques. Rather, it teaches the basics of data "cleaning", "munging" and manipulation. 

 * This book is written for aspiring data scientists, not necessarily computer scientists. Why do I draw the distinction? When discussing different types, for example, I do not discuss data structures in any depth. Rather, I discuss examples of applications where different types would be most useful. 

 * It is meant to be read on a computer with an internet connection. There are plenty of hyperlinks to click, and some interactive visualizations. These don't work if you have a text copy. 

 * This book does not attempt to be an authoritative reference. This is my attempt at balancing depth and breadth for a one-semester course. Plenty of discovery will be left to the reader. In my opinion, it is not possible to learn everything about R and Python for doing data science in one semester. The hyperlinks are there to support self-reliance and self-sufficiency. Also, I imagine it being used as a reference. It's more important that students do the exercises on their own. Associating exercises with a (sub-)section in the textbook will give them support when they need it.

 * Chapters may be read out of order, but (sub-)sections within a chapter are carefully ordered. Say, for instance, a topic in R is discussed first. If the Python discussion comes second, that discussion will reference and make comparisons with details mentioned in the R section. So read sections and subsections in order. 

## Conventions {-}

Sometimes R and Python code look very similar. At times they even look identical. This is why we usually separate R and Python code into separate sections. 

However, whenever it is necessary to prevent confusion, I will remind you what language is being used in comments (more about comments in \@ref(hello-world-in-python) ). 

```{python, collapse = TRUE}
# in python
print('hello world')
```


```{r, collapse = TRUE}
# in R
print('hello world')
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


## Installing the Required Software {-}

To get started, you must install both R and Python. The installation process depends on what kind of machine you have (e.g. what type of operating system your machine is running, is your processor $32$ or $64$ bit, etc.).

I provide some helpful links below. However, there are many alternatives available to you for how to install, and different options may be preferable to different students. Moreover, these options are likely to change over time. 

If you prefer to take a different route, or you already have a solution to either of these, feel free to keep using it. The code should still work as long as you're using relatively recent versions of each. For Python, you need a version that is `>=3.6`, and for R, it nees to be `>=4.0.0`.

### Installing R (and RStudio) {-}

It is recommended that you install R and *RStudio Desktop*. *RStudio Desktop* is a graphical user interface with many tools that making writing R easier and more fun.

Install R from the Comprehensive R Archive Network (CRAN). You can access instructions for your specific machine by clicking [here.](https://cran.r-project.org/) 

You can get RStudio Desktop directly from the [company's website](https://www.rstudio.com/products/rstudio/download/#download).

### Installing Python by Installing Anaconda {-}

It is recommended that you install *Anaconda*, which is a package manager, environment manager, and Python distribution with many third party open source packages. It provides a graphical user interface for us, too, just as RStudio does. You can access instructions for your specific machine and OS by clicking [here](https://docs.anaconda.com/anaconda/install/#).




