# Preface {-}

## About this book {-}

This book is written to be used in a one-semester statistical computing class that teaches both R and Python to graduate students in a statistics or data science department. This book is written for students that do not necessarily possess any previous familiarity with writing code. 

 * If you are using them for analyzing data, R and Python do a lot of the same things in pretty similar ways, so it does not always make sense to teach one language after the other. Imagine learning about vectorization, say, in R, and then several weeks later learning about the same concept in Python. You might end up spending a lot of time refreshing your memory before getting started for the second time. The side-by-side approach ought to help reinforce shared concepts. In my opinion, it also helps to highlight key differences.

 * This text does not describe statistical modeling techniques in detail, although many exercises will be motivated by different statistical techniques. Rather, it teaches a.) important data types, b.) the basics of common procedures such as data "cleaning", "munging" and manipulation, and c.) some background on object-oriented programming and functional programming. This is largely motivated by the setup of many graduate programs in statistics and data science. Many of these departments organize their classes by sub-discipline (e.g. time series, linear models, Bayesian, etc.). When this is the case, departments tend to compartmentalize the programming courses. The organization of this text aims to complement this kind of course offering structure.

 * This book is written for aspiring data scientists, not necessarily aspiring software developers. Why do I draw the distinction? When discussing different types, for example, I do not discuss data structures in any depth. Rather, I discuss examples of applications where different types would be most useful. 

 * This book does not attempt to be an authoritative reference. This is my attempt at balancing depth and breadth for a one-semester course. Plenty of discovery will be left to the reader. For instance, hyperlinks are provided to support self-guided exploration, and second many exercises are designed to motivate further questions. 

 * Generally speaking, chapters should be read in order. However, some jumping around may be useful. (TODO update after re-organizing) (sub-)sections within a chapter are carefully ordered. Say, for instance, a topic in R is discussed first. If the Python discussion comes second, that discussion will reference and make comparisons with details mentioned in the R section. So read sections and subsections in order. 

## Conventions {-}

Sometimes R and Python code look very similar, or even identical. This is why I usually separate R and Python code into separate sections. However, sometimes I do not, so whenever it is necessary to prevent confusion, I will remind you what language is being used in comments (more about comments in \@ref(hello-world-in-python) ). 

```{python, collapse = TRUE}
# in python
print('hello world')
```


```{r, collapse = TRUE}
# in R
print('hello world')
```



```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


## Installing the Required Software {-}

To get started, you must install both R and Python. The installation process depends on what kind of machine you have (e.g. what type of operating system your machine is running, is your processor $32$ or $64$ bit, etc.).

Below, I suggest running R with RStudio, and Python with Anaconda, and I provide some helpful links. I suggest downloading these two bundles separately; however, I should note that the recommendation below is not the only installation method. For example: 
  - one can run R and Python without downloading RStudio or Anaconda,
  - one can install RStudio with Anaconda,
  - one can run Python from within Rstudio,
  - one can run Python from within Rstudio that is managed by Anaconda, etc., and
  - options and procedures are very likely to change over time. 
  
Instructors can prefer alternative strategies, if they wish. If they do, they should verify that Python's version is `>=3.6`, and R's is `>=4.0.0`. If so, all the code in this book should run. 

### Installing R (and RStudio) {-}

It is recommended that you install R and *RStudio Desktop*. *RStudio Desktop* is a graphical user interface with many tools that making writing R easier and more fun.

Install R from the Comprehensive R Archive Network (CRAN). You can access instructions for your specific machine by clicking [here.](https://cran.r-project.org/) 

You can get RStudio Desktop directly from the [company's website](https://www.rstudio.com/products/rstudio/download/#download).

### Installing Python by Installing Anaconda {-} 

It is recommended that you install *Anaconda*, which is a package manager, environment manager, and Python distribution with many third party open source packages. It provides a graphical user interface for us, too, just as RStudio does. You can access instructions for your specific machine and OS by clicking [here](https://docs.anaconda.com/anaconda/install/#).





