--- 
title: 'An Introduction to R and Python For Data Analysis: A Side By Side Approach'
author: "Taylor R. Brown"
site: bookdown::bookdown_site
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
documentclass: krantz
lot: yes
lof: yes
fontsize: 12pt
monofont: "Source Code Pro"
monofontoptions: "Scale=0.7"
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
github-repo: tbrown122387/r_and_python_book
---

# Welcome {-}

Placeholder


## A Sample Course {-}
## License(s) {-}

<!--chapter:end:index.Rmd-->


# Preface {-}

Placeholder


## About this book {-}
## Conventions {-}
## Installing the Required Software {-}
### Installing R (and RStudio) {-}
### Installing Python by Installing Anaconda {-} 

<!--chapter:end:preface.Rmd-->


# (PART) Introducing the Basics {-} 

Placeholder


## Hello World in R
## Hello World in Python
## Getting Help
### Reading Documentation
### Understanding File Paths

<!--chapter:end:01-intro.Rmd-->


# Basic Types

Placeholder


## Basic Types In Python
### Type Conversions in Python
## Basic Types In R
### Type Conversions in R
### R's Simplification
## Exercises
### R Questions
### Python Questions

<!--chapter:end:02-basic_types.Rmd-->


# R `vector`s versus Numpy `array`s and Pandas' `Series`

Placeholder


## Overview of R
## Overview of Python
## Vectorization in R
## Vectorization in Python
## Indexing Vectors in R
## Indexing Numpy arrays
## Indexing Pandas' Series
## Some Gotchas
### Shallow versus Deep Copies
### How R and Python Handle Missing Values
## Exercises
### R Questions
### Python Questions

<!--chapter:end:03-vectors_and_arrays.Rmd-->


# Numpy `ndarray`s Versus R's `matrix` and `array` Types

Placeholder


## Numpy `ndarray`s In Python
## The `matrix` and `array` classes in R
## Exercises
### R Questions
### Python Questions

<!--chapter:end:04-arrays_and_matrices.Rmd-->


# R's `list`s Versus Python's `list`s and `dict`s

Placeholder


## `list`s In R
## `list`s In Python
## Dictionaries In Python
## Exercises
### R Questions
### Python Questions

<!--chapter:end:05-lists_and_dictionaries.Rmd-->


# Functions

Placeholder


## Defining R Functions
## Defining Python Functions
## More Details On R's User-Defined Functions
## More details on Python's user-defined functions
## Function Scope in R
## Function Scope in Python
## Modifying a Function's Arguments
### Passing By Value In R
### Passing By Assignment In Python
## Accessing and Modifying Captured Variables
### Accessing Captured Variables in R
### Accessing Captured Variables in Python
### Modifying Captured Variables In R
### Modifying Captured Variables In Python
## Exercises
### R Questions
### Python Questions

<!--chapter:end:06-functions.Rmd-->


# Categorical Data 

Placeholder


## `factor`s in R
## Two Options for Categorical Data in Pandas 
## Exercises
### R Questions
### Python Questions

<!--chapter:end:07-categorical_data.Rmd-->


# Data Frames

Placeholder


## Data Frames in R
## Data Frames in Python
## Exercises
### R Questions
### Python Questions

<!--chapter:end:08-data_frames.Rmd-->


# (PART) Common Tasks and Patterns {-} 

Placeholder


## General Input Considerations
## Reading in Text Files with R 
## Reading in Text Files with Pandas
## Saving Data in R
### Writing Out Tabular Plain Text Data in R
### Serialization in R
## Saving Data in Python
### Writing Out Tabular Plain Text Data in Python
### Serialization in Python
## Exercises
### R Questions
### Python Questions

<!--chapter:end:09-io.Rmd-->


# Using Third-Party Code

Placeholder


## Installing Packages In R 
## Installing Packages In Python 
## Loading Packages In R 
## Loading Packages In Python 
### `import`ing Examples

<!--chapter:end:10-third_party_code.Rmd-->

# Control Flow

## Conditional Logic 

We discussed Boolean/logical objects in \@ref(basic-types). We used these for

- counting the number of times a condition appeared, and 
- subsetting vectors and data frames.

Another way to use them is to conditionally execute code. You can choose to execute code depending on whether or not a Boolean/logical value is true or not. 

This is what an [if statement looks like In R](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#if): 
```{r, collapse = TRUE}
myName <- "Taylor"
if(myName != "Clare"){
    print("you are not Clare")
}
```

[In Python](https://docs.python.org/3/tutorial/controlflow.html#if-statements), you don't need curly braces, but the indentation needs to be just right, and you need a colon.

```{python, collapse = TRUE}
my_name = "Taylor"
if my_name == "Taylor":
    print("hi Taylor")
```

There can be more than one test of truth. To test alternative conditions, you can add one or more `else if` (in R) or `elif` (in Python) blocks. The first block with a Boolean that is found to be true will execute, and none of the resulting conditions will be checked.  

If no `if` block or `else if`/`elif` block executes, an `else` block will always execute. That's why `else` blocks don't need to look at a Boolean. Whether they execute only depends on the Booleans in the previous blocks. If there is no `else` block, and none of the previous conditions are true, nothing will execute.

```{r, collapse = TRUE}
food <- "muffin"
if(food == "apple"){
    print("an apple a day keeps the doctor away")
}else if(food == "muffin"){
    print("muffins have a lot of sugar in them")
}else{
    print("neither an apple nor a muffin")
}
```

```{python, collapse = TRUE}
my_num = 42.999
if my_num % 2 == 0:
    print("my_num is even")
elif my_num % 2 == 1:
    my_num += 1
    print("my_num was made even")
else:
    print("you're cheating by not using integers!")
```

## Loops

One line of code generally does one "thing," unless you're using loops. Code written inside a loop will execute many times. 

The most common loop for us will be a `for` loop. A simple [`for` loop in R](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#for) might look like this

```{r, collapse = TRUE}
myLength <- 9
r <- vector(mode = "numeric", length = myLength)
for(i in seq_len(myLength)){
    r[i] <- i
}
r
```

1. `seq_len(myLength)` gives us a `vector`.
2. `i` is a variable that takes on the values found in `seq_len(myLength)`.
3. Code inside the loop (inside the curly braces), is repeatedly executed, and it may or may not reference the dynamic variable `i`.

[Here is an example of a for loop in Python](https://docs.python.org/3/tutorial/controlflow.html#for-statements):

```{python, collapse = TRUE}
my_length = 9
r = []
for i in range(my_length):
    r.append(i)
r
```

1. Unsurprisingly, Python's syntax opts for indentation and colons instead of curly braces and parentheses.
2. Code inside the loop (indented underneath the `for` line), is repeatedly executed, and it may or may not reference the dynamic variable `i`.
3. `for` loops in Python are more flexible because they iterate over many different types of data structures--in this case [`range()` returns an object of type `range`](https://docs.python.org/3/library/stdtypes.html#range). 
4. The [`range`](https://docs.python.org/3/tutorial/controlflow.html#the-range-function) doesn't generate all the numbers in the sequence at once, so it saves on memory. This can be quite useful if you're looping over a large collection, and you don't need to store everything all at once. However, in this example, `r` is a list that *does* store all the consecutive integers.

Loops are for repeatedly executing code. `for` loops are great when you know the number of iterations needed ahead of time. If the number of iterations is not known, then you'll need a `while` loop. While loops will only terminate after a condition is found to be true. Here are some examples [in R](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#while) and [in Python](https://docs.python.org/3/reference/compound_stmts.html#while).  

```{r, collapse = TRUE, echo=FALSE}
set.seed(8)
```
```{r, collapse = TRUE}
keepGoing <- TRUE
while(keepGoing){
    oneOrZero <- rbinom(1, 1, .5)
    print(paste("oneOrZero:", oneOrZero))
    if(oneOrZero == 1)
        keepGoing <- FALSE
}
```
```{python, collapse = TRUE, echo = FALSE}
import numpy as np
np.random.seed(1)
```

```{python, collapse = TRUE}
keep_going = True
while keep_going:
    one_or_zero = np.random.binomial(1, .5) 
    print("one_or_zero: ", one_or_zero)
    if one_or_zero == 1:
        keep_going = False
```



::: {.rmd-caution}
Here are some tips for writing loops:

1. If you find yourself copying and pasting code, changing only a small portion of text on each line of code, you should consider using a loop. 
2. If a `for` loop works for something you are trying to do, first try to find a replacement function that does what you want. The examples above just made a `vector`/`list` of consecutive integers. There are many built in functions that accomplish this. Avoiding loops in this case would make your program shorter, easier to read, and (potentially) much faster. 
3. A third option between looping, and a built-in function, is to try the functional approach. This will be explained more in the last chapter. 
3. Watch out for [**off-by-one** errors](https://en.wikipedia.org/wiki/Off-by-one_error). Iterating over the wrong sequence is a common mistake, considering 
    * Python starts counting from $0$, while R starts counting from $1$
    * sometimes iteration `i` references the `i-1`th element of a container
    * The behavior of loops is sometimes more difficult to understand if they're using [`break`](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Looping) or [`continue`/`next` statements](https://docs.python.org/3/tutorial/controlflow.html#break-and-continue-statements-and-else-clauses-on-loops).
4. *Don't hardcode variables.* In other words, don't write code that is specific to particulars of your script's current draft. Write code that will still run if your program is fed different data, or if you need to calculate something else that's closely-related (e.g. run the same calculations on different data sets, or vary the number of simulations, or make the same plot several times in similar situations, etc.). I can guarantee that most of the code you write will need to be run in many different situations. If, at every time you decide to make a change, you need to hunt down multiple places and make multiple changes, there is a nontrivial probability you will miss at least one. As a result, you will introduce a bug into your program, and waste (sometimes a lot of) time trying to find it.
5. Watch out for infinite `while` loops. Make sure that your stopping criterion is guaranteed to eventually become true.
:::

<!-- Above I provide some advice that is worth repeating: *don't hardcode variables*. The goal is to minimize the number of places you have to make changes to your code, if such a need arises. , so try to save your future self some time.  -->

<!-- For example, in statistical programs, there are often "tuning parameters" that must be changed frequently to affect the overall behavior of the program. For instance, you might need to re-run some plotting script after changing a few graphical parameters, or you might need to change the confidence or significance level of some interval or hypothesis test, or change a variable you're using to subset values in a data frame, or alter the number of simulations you're performing, or the number of iterations you're performing, etc. If these variables only need to be changed in one location, that saves you a lot of time and gives you more flexibility. If they don't need to be changed at all, even better!  -->

<!-- On the other hand, if you refer to specific choices in the code of your program, then every time you make a change, you will need to hunt down all of those lines of code, and make multiple changes. -->

<!-- In the code examples above, the  `myLength` and `my_length` variables could be referenced in many places later on in our programs.  -->

<!-- Even better, use a function to calculate the length of the container. Then, even if your container changes in size, the program will still run as intended.  -->

<!-- TODO -->
Python also provides an alternative way to construct lists similar to the one we constructed in the above example. They are called [**list comprehensions**](https://docs.python.org/3/tutorial/datastructures.html#list-comprehensions). These are convenient because you can incorporate iteration and conditional logic in one line of code. 

```{python, collapse = TRUE}
[3*i for i in range(10) if i%2 == 0]
```
You might also have a look at [*generator expressions*](https://www.python.org/dev/peps/pep-0289/) and [*dictionary comprehensions*](https://www.python.org/dev/peps/pep-0274/).

R can come close to replicating the above behavior with vectorization, but the conditional part is hard to achieve without subsetting.

```{r, collapse = TRUE}
3*seq(0,9)[seq(0,9)%%2 == 0]
```




## Exercises

<!--chapter:end:11-control_flow.Rmd-->


# Reshaping and Combining Data Sets

Placeholder


## Ordering and Sorting Data
## Stacking Data Sets and Placing Them Shoulder to Shoulder
## Merging or Joining Data Sets
## Long Versus Wide Data
### Long Versus Wide in R
### Long Versus Wide in Python

<!--chapter:end:12-reshaping.Rmd-->


# Visualization

Placeholder


## Base R Plotting
## Plotting with `ggplot2`
## Plotting with Matplotlib

<!--chapter:end:13-visualization.Rmd-->

# Working With Text Data

TODO

- regular expressions


<!--chapter:end:14-text_data.Rmd-->

# Dates and Times

TODO


<!--chapter:end:15-dates_and_times.Rmd-->

# Running Scripts from the Command Line

TODO


<!--chapter:end:16-CLI.Rmd-->


# (PART) Programming Styles {-} 

Placeholder


## OOP In Python
### Overview
### A First Example
### Adding Inheritance
### Adding in Composition
## OOP In R
### S3 objects: The Big Picture
### Using S3 objects
### Creating S3 objects
### S4 objects: The Big Picture
### Using S4 objects
### Creating S4 objects
### Reference Classes: The Big Picture
### Creating Reference Classes
### Creating R6 Classes
## Exercises

<!--chapter:end:17-oop.Rmd-->


# Functional Programming

Placeholder


## Functions as Function Inputs in R
### `sapply()` and `vapply()`
### `lapply()`
### `apply()`
### `tapply()`
### `mapply()`
### `Reduce()` and `do.call()`
## Another Example in R
## Functions as Function Inputs in Base Python
### `map()`
### `filter()`
## Functions as Function Inputs in Numpy
## Functional Methods in Pandas
## Functions as Function Inputs (miscellany)
## Functions as Function Outputs in R
## Functions as Function Outputs in Python

<!--chapter:end:18-FP.Rmd-->

